Metacircular Virtualization \& \\ Practical Nock Interpretation

### 2025-08-28 mss.tex
Post

Metacircularity was considered sufficiently important that several programming languages implemented it as a first-class feature, including Forth, TeX, and Prolog.  They have generally been used either for tightly interpreted features like debuggers and code introspection tools, or for language extensions like macros.

@@

\noindent
For all the mechanics involved in opcode 9, the implementation is perhaps surprisingly simple.  Extract the arm at the axis and evaluate it structurally using an opcode 2.  (As an aside, Hoon which compiles to Nock aggressively utilizes lambdas to define functions at the point of use, and frequently employs a gate-building gate pattern.  This latter resolves to two opcode 9s in a series.)

@@

Opcode 12 is the scry function, which retrieves a value from the \lstinline[style=inlinecode]{sky} namespace environment.\footnote{There is a sense in which opcodes 6--11 are also virtualized extensions to Nock, acting as they do as macros.}  This is implemented in \lstinline[style=inlinecode]{++mink} as follows:

@@

\sloppy
\lstinline[style=inlinecode]{++mink} is naturally jetted for efficiency, ultimately by \lstinline[style=inlinecode]{u3m_soft_run()} in Urbit's Vere (C) and by \lstinline[style=inlinecode]{interpret()} in NockApp's NockVM (n\'{e}e) Sword (Rust), both the virtualization context for their respective runtimes.  Some internal affordances for the runtime are in place to keep execution consistent; e.g., scry gates are stored on a stack to prevent re-entry if it scries.

>>>>

Metacircularity was considered sufficiently important that several programming languages implemented it as a first-class feature, including Forth, \TeX, and Prolog.  They have generally been used either for tightly interpreted features like debuggers and code introspection tools, or for language extensions like macros.

@@

\noindent
For all the mechanics involved in opcode 9, the implementation is perhaps surprisingly simple.  Extract the arm at the axis and evaluate it structurally using an opcode 2.  (As an aside, Hoon which compiles to Nock aggressively utilizes lambdas to define functions at the point of use, and frequently employs a gate-building gate pattern.  This latter resolves to two opcodes 9 in series.)

@@

Opcode 12 is the scry function, which retrieves a value from the \lstinline[style=inlinecode]{sky} namespace environment.\footnote{There is a sense in which opcodes 6--11 are also virtualized extensions to Nock, acting as they do as macros.  See \patp{niblyx-malnus}, pp.~47–70 in this volume.}  This is implemented in \lstinline[style=inlinecode]{++mink} as follows:

@@

\sloppy
\lstinline[style=inlinecode]{++mink} is naturally jetted for efficiency, ultimately by \lstinline[style=inlinecode]{u3m_soft_run()} in Urbit's Vere (C) and by \lstinline[style=inlinecode]{interpret()} in NockApp's NockVM (formerly Sword) (Rust), both the virtualization context for their respective runtimes.  Some internal affordances for the runtime are in place to keep execution consistent; e.g., scry gates are stored on a stack to prevent re-entry if it scries.
